
\documentclass{article}[12pt]

\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{graphicx}
\usepackage{cleveref}
\usepackage{natbib}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{array}
\usepackage{float}
\usepackage{caption}

\captionsetup[table]{width=0.95\textwidth, font=small}
\captionsetup[figure]{width=0.95\textwidth, font=small}

\hypersetup{
	colorlinks=true,
	linkcolor=black,
	citecolor=black}

%%% TIKZ graphics
\usetikzlibrary{positioning, calc, shapes.geometric, shapes.multipart, shapes, arrows.meta, arrows, 	decorations.markings, external, trees}

\tikzstyle{Arrow} = [
thick,
decoration={
	markings,
	mark=at position 0.99 with {  % fix bug with tikz arrow head interference with decoration
		\arrow[thick]{latex}
	}
},
shorten >= 3pt, preaction = {decorate}
]
\tikzstyle{BiArrow} = [
thick,
decoration={
	markings,
	mark=at position 0.99 with {  % fix bug with tikz arrow head interference with decoration
		\arrow[thick]{latex}
	}
},
shorten >= 3pt, preaction = {decorate}
]
\tikzstyle{RedArrow} = [
red, 
line width = 0.5mm,
decoration={
	markings,
	mark=at position 0.99 with {
		\arrow[line width = 0.5mm,red]{latex}
	}
},
shorten >= 3pt, preaction = {decorate}
]

\tikzstyle{BlueArrow} = [
blue, 
line width = 0.5mm,
decoration={
	markings,
	mark=at position 0.99 with {
		\arrow[line width = 0.5mm,blue]{latex}
	}
},
shorten >= 3pt, preaction = {decorate}
]

\tikzstyle{Dashed} = [
thick,dashed,
decoration={
	markings,
	mark=at position 0.99 with {  % fix bug with tikz arrow head interference with decoration
		\arrow[thick]{latex}
	}
},
shorten >= 3pt, preaction = {decorate}
]

\tikzset{
    BiArrow/.style={
        thick,
        decoration={
            markings,
            mark=at position 0.99 with {
                \arrow[thick]{latex}
            },
            mark=at position 0 with {
                \arrow[thick]{latex}
            }
        },
        shorten >= 3pt,
        preaction = {decorate}
    }
}



\newtheorem{remark}{Remark}
\AfterEndEnvironment{remark}{\noindent\ignorespaces}
\newtheorem{assumption}{Assumption}
\AfterEndEnvironment{assumption}{\noindent\ignorespaces}
\newtheorem{proposition}{Proposition}
\AfterEndEnvironment{proposition}{\noindent\ignorespaces}
\newtheorem{theorem}{Theorem}
\AfterEndEnvironment{theorem}{\noindent\ignorespaces}
\newtheorem{corollary}{Corollary}
\AfterEndEnvironment{corollary}{\noindent\ignorespaces}
\newtheorem{condition}{Condition}
\AfterEndEnvironment{condition}{\noindent\ignorespaces}
\newtheorem{lemma}{Lemma}
\AfterEndEnvironment{lemma}{\noindent\ignorespaces}

% \AtAppendix{\counterwithin{lemma}{section}}

\theoremstyle{definition}
\newtheorem{exmp}{Example}
\AfterEndEnvironment{exmp}{\noindent\ignorespaces}
\newtheorem{design}{Design}
\AfterEndEnvironment{design}{\noindent\ignorespaces}
\newtheorem{defi}{Definition}
\AfterEndEnvironment{defi}{\noindent\ignorespaces}

\def \bc {\boldsymbol{c}}
\def \bz {\boldsymbol{z}}
\def \bZ {\boldsymbol{Z}}
\def \bA {\boldsymbol{A}}
\def \ba {\boldsymbol{a}}
\def \bX {\boldsymbol{X}}
\def \bx {\boldsymbol{x}}
\def \bY {\boldsymbol{Y}}
\def \wY {\widetilde{Y}}
\def \bI {\boldsymbol{I}}
\def \bS {\boldsymbol{S}}
\def \bR {\boldsymbol{R}}
\def \bO {\boldsymbol{O}}
\def \bG {\boldsymbol{G}}
\def \bE {\boldsymbol{E}}
\def \bV {\boldsymbol{V}}
\def \bf {\boldsymbol{f}}
\def \bD {\boldsymbol{D}}
\def \cR {\mathcal{R}}

\def \EX {\mathbb{E}}

\def \beps {\boldsymbol{\varepsilon}}

\newcommand{\indep}{\mathrel{\perp\!\!\!\perp}}
\newcommand{\cindep}[3]{#1 \indep #2 \mid #3}

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%


\usepackage[status=draft,inline,index]{fixme}
%\usepackage[status=final,inline,index]{fixme}
\fxsetup{theme=color,margin=false,mode=multiuser}
\FXRegisterAuthor{bw}{bwe}{BW}


\title{Sensitivity analysis for contamination in cluster randomized trials with interference}

\author{Bar Weinstein}
% date is current date
\date{\today}

\begin{document}
\maketitle


\section{Problem setup}
Consider a CRT with one stage randomization. Assume there are clusters
$C_1, \ldots, C_K$ in the population, where $K$ is the total number of clusters.
Let $n_k$ be the size of cluster $C_k$, and $N = \sum_{k=1}^K n_k$ be the total population size.
Assume researchers randomize binary treatments at the cluster level. 
Let $Z_k$ be the treatment assignment of cluster $C_k$, where $Z_k=1$ if cluster $C_k$ is treated, and $0$ otherwise.
Assume that $P(Z_k = 1) = p_z \in (0,1)$ for all $k=1,\ldots,K$.

The observed network $\widetilde{\bA}$ consists of $K$ disjoint clusters, each 
assumed to be fully connected. The true network $\bA$ may contain edges between clusters,
leading to contamination across clusters.

Let $Z_{ik}$ be the treatment assignment of unit $i$ in cluster $C_k$.
Assume the binary exposure mapping
\begin{assumption}[Exposure mapping]
\label{ass:expos_map}
    Let $F(\bZ_{-i},\bA) =\mathbb{I}\left\{\sum_{j\neq i} Z_jA_{ij} > 0\right\}$. For any $\bz, \bz' \in \mathcal{Z}$, if $z_i = z'_i$ and $F(\bz_{-i},\bA) = F(\bz'_{-i},\bA)$, then $Y_i(\bz) = Y_i(\bz')$, 
\end{assumption}
where $\mathcal{Z} \subseteq \{0,1\}^N$ is the space of all possible treatments allocations with 
non-zero probability.
onsequently, each unit has four effective potential outcomes,
$Y_i(z,f)$ for $z,f\in\{0,1\}$, indicating whether unit $i$ is treated and whether at least one of its neighbors is treated.
The observed exposures are $\widetilde{F}_i = F(\bZ_{-i}, \widetilde{A})$.
By the CRT design, $\widetilde{F}_i = Z_i$ with probability one, 
as units within a cluster share the same treatment assignment, and the observed network 
consists of disjoint clusters.

Assume that the observed outcomes $Y_i$ are generated from one of the potential outcomes by:
\begin{assumption}[Consistency]
\label{ass:consis}
$Y_i = \sum_{z,f \in \{0,1\}}Y_i(z,f) \mathbb{I}\left\{Z_i=z, F_i=f\right\}$ .    
\end{assumption}

The causal estimand is the sample average treatment effects. Using our setup and notations is it reduced to
\begin{equation}
    \label{eq:ate}
    ATE = \frac{1}{N}\sum_{i=1}^N Y_i(1, 1) - Y_i(0, 0) = 
    \frac{1}{N}\sum_{k=1}^{K}\sum_{i \in C_k} Y_i(1, 1) - Y_i(0, 0).
\end{equation}
The most common estimator for the ATE in CRTs is the Horvitz-Thompson (HT) estimator,
\begin{equation}
    \label{eq:ate_ht}
    \begin{aligned}
        \widehat{ATE} 
        &= 
        \frac{1}{N}
        \sum_{k=1}^{K}
        \sum_{i \in C_k}
        \left[
        \frac{
        \mathbb{I}\{Z_k=1\}Y_i}{p_z} 
        -
        \frac{
        \mathbb{I}\{Z_k=0\}Y_i}{1 - p_z} 
        \right]
        \\ &=
        \frac{1}{N}
        \sum_{i=1}^{N}
        \left[
        \frac{  
        \mathbb{I}\{Z_i=1\}Y_i}{p_z}
        -
        \frac{
        \mathbb{I}\{Z_i=0\}Y_i}{1 - p_z}
        \right].
    \end{aligned}
\end{equation}
For units in a treated cluster, the observed exposure $\widetilde{F}_i=1$ agrees with the true exposure $F_i=1$.
Therefore
\begin{equation*}
    \EX_{\bZ} \left[\frac{
        \mathbb{I}\{Z_i=1\}Y_i}{p_z}
    \right]
    = Y_i(1,1).
\end{equation*}
However, for units in a control cluster, the observed exposure $\widetilde{F}_i=0$ 
may not agree with the true exposure $F_i$, which can be either $0$ or $1$ due to contamination.
Notice that $\mathbb{I}\{Z_i=0\} = \mathbb{I}\{\widetilde{F}_i=0\}$
with probability one as $\widetilde{F}_i=1$ only if $Z_i=1$.
Thus, 
\begin{equation*}
    \begin{aligned}
        \EX_{\bZ} \left[\frac{
            \mathbb{I}\{Z_i=0\}Y_i}{1 - p_z}
        \right]
        &= 
        \EX_{\bZ} \left[\frac{
            \mathbb{I}\{\widetilde{F}_i=0\}Y_i}{1 - p_z}
        \right]
        \\ &=
        \EX_{\bZ} \left[\frac{
            \mathbb{I}\{\widetilde{F}_i=0, F_i=0\}Y_i(0,0)}{1 - p_z}
        \right]
        \\&\quad +
        \EX_{\bZ} \left[\frac{
            \mathbb{I}\{\widetilde{F}_i=0, F_i=1\}Y_i(0,1)}{1 - p_z}
        \right]
        \\ &=
        Y_i(0,0) \Pr(F_i = 0 \mid \widetilde{F}_i=0)
        \\&\quad +
        Y_i(0,1) \Pr(F_i = 1 \mid \widetilde{F}_i=0).
    \end{aligned}
\end{equation*}
By Bayes' rule,
\begin{equation*}
    \begin{aligned}
        \Pr(F_i = 0 \mid \widetilde{F}_i=0) 
        &=
        \frac{
            \Pr(\widetilde{F}_i=0 \mid F_i=0) \Pr(F_i=0)
        }{
        \Pr(\widetilde{F}_i=0)
        }
        \\ &=
        \frac{
            1-\pi_i
        }{
          1-p_z
        }.
    \end{aligned}
\end{equation*}
Where we used the fact that $ \Pr(\widetilde{F}_i=0 \mid F_i=0) = 1$
as $\sum_{j\neq i}\widetilde{A}_{ij}Z_j \leq \sum_{j\neq i}A_{ij}Z_j$ with 
probability one. Therefore, $\widetilde{F}_i \leq F_i$ as well. Since 
the exposures are binary, $F_i=0$ implies $\widetilde{F}_i=0$ with probability one,
Also, $\Pr(\widetilde{F}_i=0) = \Pr(Z_i=0) = 1-p_z$ by the CRT design.
And we use the notation $\pi_i = \Pr(F_i=1)$.
Consequently,
\begin{equation*}
     \EX_{\bZ} \left[\frac{
            \mathbb{I}\{Z_i=0\}Y_i}{1 - p_z}
        \right] 
        = 
        \frac{
            (1- \pi_i)Y_i(0,0) + (\pi_i-p_z)Y_i(0,1)
        }{
            1-p_z
        }.
\end{equation*}
That enables us to express the expectation of the HT estimator as
\begin{equation*}
    \begin{aligned}
        \EX_{\bZ}[\widehat{ATE}] 
        &=
        \frac{1}{N}
        \sum_{i=1}^{N}
        \left[
        Y_i(1,1)
        -
        \frac{
            (1- \pi_i)Y_i(0,0) + (\pi_i-p_z)Y_i(0,1)
        }{
            1-p_z
        }
        \right]
        \\ &=
        ATE
        -
        \frac{1}{N}
        \sum_{i=1}^{N}
        \frac{
            \pi_i-p_z 
        }{
            1-p_z
        }
        \left(Y_i(0,1) - Y_i(0,0)\right).
    \end{aligned}
\end{equation*}
Notice that 
\begin{equation*}
    \begin{aligned}
        \pi_i &=
        \Pr\left(F_i=1\right)
        \\ &= 
        \Pr\left(F_i = 1 \mid Z_i=1\right) \Pr(Z_i=1)
        + \Pr\left(F_i = 1 \mid Z_i=0\right) \Pr(Z_i=0)
        \\ &=
        p_z + (1-p_z) \Pr\left(F_i = 1 \mid Z_i=0\right)
        \\ & \geq p_z.
    \end{aligned}
\end{equation*}
Therefore, if $Y_i(0,1) \geq Y_i(0,0)$ for all $i$,
then $\EX_{\bZ}[\widehat{ATE}] \leq ATE$, and if 
$Y_i(0,1) \leq Y_i(0,0)$ for all $i$,
then $\EX_{\bZ}[\widehat{ATE}] \geq ATE$.

\subsection*{Bias-corrected estimator}
Similarly to the direct effects in the ENRT design,
we don't have any information about $Y_i(0,1)$ in the observed data.
We have to make an additional structrual assumption to proceed. 
One approach is to assume 
\begin{assumption}
    \label{ass:po_prop}
    There exists a constant $\delta$ such that for all $i=1,\ldots,N$,
    \begin{equation*}
        Y_i(0,1) - Y_i(0,0) = 
    \delta\left\{
        Y_i(1,1) - Y_i(0,0).
    \right\}
    \end{equation*} 
\end{assumption}
The sensitivity parameter can be interpreted as the ratio between indirect 
effects (given no direct treatment) to the overall or average effects
$\delta = \frac{IE_i}{ATE_i}$:
\begin{itemize}
    \item $\delta=0$: no indirect effects. Contamination from other clusters 
                   has no different effect.
    \item $\delta=1$: indirect effects are equal to the overall effects.
             Direct treatment $Z_i=1$ has no additional benefit beyond indirect exposure.
   \item $\delta \in (0,1)$. Indirect effect is positive but smaller than the total effect.
\end{itemize}
We can decompose 
\begin{equation*}
   \begin{aligned}
    ATE_i &= Y_i(1,1) - Y_i(0,0) 
    \\ &= \left\{Y_i(1,1) - Y_i(0,1) \right\} + \left\{Y_i(0,1) - Y_i(0,0) \right\}
    \\ &= DE_i + IE_i,
   \end{aligned}
\end{equation*}
thus, under Assumption~\ref{ass:po_prop}, 
\begin{equation*}
        ATE_i = DE_i + \delta ATE_i \Rightarrow DE_i = (1-\delta) ATE_i.
\end{equation*}
So assumptions about $\delta$ are also about the decomposition of 
the overall ATE into a direct and indirect parts. 
We didn't limit $\delta \in [0,1]$ in Assumption~\ref{ass:po_prop},
which in theory allows extrapolations. This can modified accordingly.

By Assumption~\ref{ass:po_prop}, $Y_i(0,1) = \delta Y_i(1,1) + (1-\delta)Y_i(0,0)$.
Substituting it into the expected value of the naive estimator yield
\begin{equation*}
    \begin{aligned}
        \EX_{\bZ}[\widehat{ATE}] &=
         ATE
        -
        \frac{1}{N}
        \sum_{i=1}^{N}
        \frac{
            \pi_i-p_z 
        }{
            1-p_z
        }
        \left(\delta Y_i(1,1) + (1-\delta)Y_i(0,0)
         - Y_i(0,0)\right)
         \\&= 
         ATE
         -
         \frac{1}{N}
        \sum_{i=1}^{N}
        \frac{
            \delta(\pi_i-p_z)
        }{
            1-p_z
        }
        \left(Y_i(1,1) - Y_i(0,0)\right)
        \\ &=
        \frac{1}{N}
        \sum_{i=1}^{N}
        \left[
         1 - 
       \frac{
            \delta(\pi_i-p_z)
        }{
            1-p_z
        }
        \left(Y_i(1,1) - Y_i(0,0)\right)  
        \right] 
        \\ &=
        \frac{1}{N}
        \sum_{i=1}^{N}
        \frac{
            1 - p_z(1-\delta) - \delta \pi_i
        }{
            1-p_z
        }
        \left(Y_i(1,1) - Y_i(0,0)\right).
    \end{aligned}
\end{equation*}
Which motivates the bias-corrected estimator
\begin{equation}
    \label{eq:ate_ht_crt_bc}
        \widehat{ATE}_{adj} =
        \frac{1}{N}
        \sum_{i=1}^{N}
        \frac{1 - p_z}{
             1 - p_z(1-\delta) - \delta \pi_i
        }
        \left[
        \frac{  
        \mathbb{I}\{Z_i=1\}Y_i}{p_z}
        -
        \frac{
        \mathbb{I}\{Z_i=0\}Y_i}{1 - p_z}
        \right].
\end{equation}
Which satisfies $\EX_{\bZ}[\widehat{ATE}_{adj}] = ATE$.
Notice that the case of $\delta=1$ recovers the bias-corrected indirect 
effect estimator in the ENRT design.

Eq.~\eqref{eq:ate_ht_crt_bc} motivates a two-parameter sensitivity analysis
for contamination in CRTs. The parameters are $\delta$ and $\pi_i$.
Under conditional edge independence assumptions, we can express $\pi_i$ as a function of
the individual-level contamination probabilities (see the ENRT paper).
Such specification can take into account prior knowledge about 
contamination from which clusters to which clusters is feasible.
For example, contamination between geographically distant clusters is less likely,
or contamination between hospitals in different health networks is not possible.
For $\delta$, I think it is reasonable to limit it to $[0,1]$ in practice,
as discussed above.

Extending \eqref{eq:ate_ht_crt_bc} to include an outcome model 
$\mu_i(z) = \EX[Y_i \mid Z_i=z, \bX_i]$ is straightforward,
and be done as in the ENRT paper.
Similarly, the variance estimators derived in the ENRT paper 
for the indirect effect estimator is cluster-robust and can be adapted to the CRT case.

\end{document}
